---
title: "Project---Nutrition"
author: "Yutong Liu  Lin Zhao"
date: "11/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
```

## Data exploration

```{r cars}
library(dplyr)
library(tidyr)
food_component <- read.csv("food_component.csv")
food_nutrient <- read.csv("food_nutrient.csv")
nutrient <- read.csv("nutrient.csv")
food_component_nutrient <- inner_join(food_component,food_nutrient,by = "fdc_id") # This results in 6300 observarions of 18 variables
food_component_nutrient <- inner_join(food_component_nutrient, nutrient, by= c("nutrient_id"="id") ) # add the nutrient name instead of nutrient id
food_dataset <- food_component_nutrient[,c(3,4,5,6,11,14,15,16,19,20)] # the "food nutrient of food component" analysis
col_names <- c("component_name","pct_weight","is_refuse","gram_weight","nutrient_amount","min","max","median","nutrient_name","nutrient_unit")
colnames(food_dataset) <- col_names
food_dataset <- food_dataset[,c(1,2,3,4,9,5,6,7,8,10)]
food_dataset <- food_dataset %>% mutate(nutrition_amount_in_component= nutrient_amount/100*gram_weight)  
```

Calculate the exact nutrition amount in each component.

```{r}
library(ggplot2)
ggplot(food_dataset)+ geom_bar (aes(x= nutrient_unit, fill = nutrient_unit, weight = nutrition_amount_in_component))
```

> Explain what this graph shows before you move on to another graph!

```{r}
ggplot(food_dataset)+ geom_bar (aes(x= is_refuse, fill =component_name, weight = nutrition_amount_in_component))
```

> More explanation, maybe even motivate the next graph before you make it.

```{r}
food_dataset <- food_dataset %>% filter(is_refuse == 'Y') %>% group_by (nutrient_name)
ggplot(food_dataset)+ geom_bar(aes(x= component_name, fill =component_name, weight = nutrition_amount_in_component)) + coord_flip() + facet_wrap(~nutrient_unit)
```

> Note that using the weight argument results in *dramatically* different charts.
Making the plot to look at the nutrient unit, finding the better way to do the unit adjustment.

> Also, you should be explaining each graph. Tell a story with your data! What are you supposed to be seeing?

Looking at the data composition by categorizing the refused and non-refused food components, and the specific component occupation ratio in these two categories, and then filtering the refused food component for further analysis


## Unit Adjustment

Mass unit adjustment: 1 g = 1000 mg = 1000000 ug

Energy unit adjustment: 1 KCAL = 4.184 KJ

```{r}
if ( food_dataset$nutrient_unit == 'MG') {
  food_dataset$nutrition_amount_in_component = food_dataset$nutrition_amount_in_component/ 1000
} else if ( food_dataset$nutrient_unit == 'UG') { 
  food_dataset$nutrition_amount_in_component = food_dataset$nutrition_amount_in_component/1000000
} else if ( food_dataset$nutrient_unit == 'KCAL') {
  food_dataset$nutrition_amount_in_component = food_dataset$nutrition_amount_in_component* 4.184
}
```

Adjusting the mass unit to G, the energy unit to KJ, therefore, there exists mass unit-G, energy unit-KJ and vitamin unit-IU. 


```{r}
unit<- sub("MG",  "G", food_dataset$nutrient_unit) 
unit<- sub("UG",  "G", unit)
unit<- sub("KCAL",  "kJ", unit)
food_dataset <- food_dataset %>% replace(., "nutrient_unit", c(unit)) %>% unite(., "nutrient_name_nutrient_unit", nutrient_name, nutrient_unit)
```

Combine the nutrient name and unit together, so when the audience use the interactive graphics, they will know the exact quantity of nutrient and compare the nutrient amount in different component.

The interactive graphics saved in 'explore_table.R' file show the total and weighted nutrient amount for each component.

We take Vitamin E as an example, create vitamin E dataset, looking for which component is best for extract vitamin E.

```{r}
library(tidyverse)
vitaminE <- food_dataset %>% filter(str_detect(nutrient_name_nutrient_unit,"Vitamin E")) %>% 
  group_by(component_name) %>% 
  summarize(pct_weight = sum(pct_weight),  gram_weight = sum(gram_weight),
            nutrient_amount = sum(nutrient_amount),
            min = sum(min), max = sum(max), median = sum(median),
            nutrition_amount_in_component = sum(nutrition_amount_in_component))
ggplot(vitaminE) + 
  geom_bar (aes(x= component_name, y = nutrient_amount/gram_weight), stat = "identity") +
  theme(axis.text.x = element_text(angle=45, size = 8))

```




