---
title: "Stat 850 Project Description"
author: "Constanza Paz Avello Lefno, Yutong Liu, Lin Zhao"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 3)
```

```{r load-sas-libraries, echo = F, message = F, warning = F}

saspath <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en" # Your SAS path goes here
sasopts <- "-ls 75"
knitr::opts_chunk$set(engine.path=list(sas=saspath, saslog=saspath, sashtml=saspath),
                      engine.opts=list(sas=sasopts, saslog=sasopts, sashtml=sasopts), 
                      comment=NA)
library(SASmarkdown)
```

## Instructions

Each member of your team should modify this document in some way and push their modifications to the repository in a separate commit. This will ensure that you have set your repository up in a way that ensures all group members are working with the same repository. 

Note that you must compile your readme (this document) for it to properly display as part of your github repository. 

Once you have received feedback on your project proposal (via Canvas) you may alter this README so that it describes your final project instead of the project proposal. 

## Data Set

[FoodData](https://fdc.nal.usda.gov/fdc-datasets/FoodData_Central_csv_2020-04-29.zip) is an integrated dataset that provides food component and nutrient information.

Three tables from the dataset list are chosen for this project:

* food_component.csv
* food_nutrient.csv
* nutrient.csv

By combining three datasets together and selecting several variables, a new dataset called **food_dataset** is generated for further analysis.

Description of variables:

* component_name
  
   _The kind of component, e.g. bone_
   
* pct_weight
  
   _The weight of the component as a percentage of the total weight of the food_
   
* is_refuse
  
   _Whether the component is refuse, i.e. not edible_

* gram_weight
  
   _The weight of the component in grams_

* nutrient_amount
  
   _Amount of the nutrient per 100g of food. Specified in unit defined in the nutrient table._
   
* min
  
   _The minimum amount_

* max
  
   _The maximum amount_

* median
  
   _The median amount_

* nutrient_name
  
   _Name of the nutrient_

* nutrient_unit
  
   _The standard unit of measure for the nutrient (per 100g of food)_


## Potential Topics to Explore Using the Data Set
Describe avenues you might explore using the data

## Group Members

* Constanza Paz Avello Lefno
* Yutong Liu
* Lin Zhao

## Reference

U.S. Department of Agriculture, Agricultural Research Service. FoodData Central, 2019. fdc.nal.usda.gov.
